Drought regimens predict life history strategies in \emph{Heliophila}

### 2019-03-31 manuscript/manuscript.tex
revised

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Drought frequency predicts life history strategies in Heliophila},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn G. Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records, Heliophia},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Drought frequency predicts life history strategies in \emph{Heliophila}}
    \author{J. Grey Monroe\textsuperscript{1,2}, Brian Gill\textsuperscript{3},
Kathryn G. Turner\textsuperscript{4}, \& John K.
McKay\textsuperscript{2}}
    \date{}

\shorttitle{Drought and life history}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO 80521, USA\\\textsuperscript{2} College of Agriculture, Colorado State University, Fort Collins, CO 80521, USA\\\textsuperscript{3} Institute for Environment and Society, Brown University, Providence, RI 02912, USA\\\textsuperscript{4} Biology Department, Pennsylvania State University, State College, PA 16802, USA}
\keywords{drought adaptation, life history evolution, remote sensing, phylogeography, herbaria records, Heliophia}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

Explaining variation in life history strategies is an enduring goal of
evolutionary biology and ecology. For plants, annual and perennial life
histories have long been thought to reflect adaptation to environments
that experience alternative drought regimes. Neverthless, empirical
support for this hypothesis from phylogeographic analyses remains
lacking.

@@

Here, we tested classic life history theory in \textit{Heliophila}
(Brassicaceae), a diverse genus of flowering plants native to Africa,
with a phylogentically constrained analysis integrating 34 years of
satellite-based drought detection with 2192 herbaria occurrence records.

@@

We found that annual and perennial life histories have evolved multiple
times in \textit{Heliophila} and that compared to annuals, perennial
species occur in environments where droughts are significantly less
frequent. We also report evidence that annual plants adapt to
predictable drought regimes by escaping drought prone seasons as seeds.

@@

These results provide compelling support for longstanding theoretical
predictions and a more nuanced understanding of life history evolution.
More broadly, this work outlines a scalable framework integrating
herbaria records and remote sensing to address outstanding questions in
evolutionary ecology. 
}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with some species
completing their life cycle in a number of weeks to others that live for
thousands of years (Brown, 1996). Along this continuum in angiosperms an
important division exists between annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons
but can also benefit from multiple bouts of reproduction and competative
advantages. These represent fundamentally different life history
strategies and predicting their occurance is important for community,
ecosystem, and agricultural ecology. However, the environmental factors
that explain their evolution and distributions remain empirically
uresolved (Friedman \& Rubin, 2015).

Classical theory predicts shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Stearns, 1992; Franco \&
Silvertown, 1996). Because lack of water is perhaps the greatest threat
to survival during vegetative or reproductive growth in plants, this
theory has been extended to the hypothesis that annuality is adaptive
when it allows plants to escape drought (Schaffer \& Gadgil, 1975).
Indeed, adaptation to drought, defined as episodes of increased aridity
causing plant sresss (Passioura, 1996), is often invoked as an
explanation for the success of annual species. And while a few cases are
cited where annuality appears to be more common in environments with
greater aridity (Stebbins Jr, 1952; Morishima \emph{et al.}, 1984), this
hypothesis has yet to be supported in a phylogenetic context. In one
previous study where this question was addressed phylogenetically,
(Evans \emph{et al.}, 2005) annuals were not found to be associated with
environments that experience more drought. This could be explained by
the relatively small number of species studied and the reliance on a
limited number of weather stations to characterize environments,
highlighting the need to develop more scalable methods to study trait
evolution phylogeographically.

It is also critical to consider another dimension of drought adaptation:
the expectation that annuality is most adaptive when droughts are not
only frequent but also predictable. That is, when the frequency of
drought is particularly high during certain seasons. Such predictability
is important for selection to lead to escape from those seasons which
are paricuarlary drought prone. While there has been at least one
example of annuality associated with environments qualitatively
classified as \enquote{predictable} in a general sense (Datson \emph{et
al.}, 2008), the seasonal predictability of drought experienced by
annuals has yet to be rigorously studied. As such, further empirical
work is needed to support the model of annuality as a mechanism of
drought adaptation via escape from drought prone seasons.

In addition to drought escape in annuals as a mechanism of adaptation to
frequent predictable droughts, droughts may be necessary for the success
of annuals more generally by acting as episodes of disturbance that
provide opportunites for annuals to establish and compete with sympatric
perennial species. Indeed, there is considerable evidence that
perennials dominate in envirnoments where distrurbance events are
infrequent (Rees \& Long, 1992; Corbin \& D'Antonio, 2004; Clary, 2012).
The resulting prediction from this hypothesis is that in the absence of
frequent drought, perenniality should evolve. However, little is known
about this component of life history evolution because previous work has
largely focused on the origins of annuality rather than perenniality
(Friedman \& Rubin, 2015).

Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses within the African endemic mustard genus,
\emph{Heliophila} L. (Brassicaceae). If annuality is an adaptive
strategy allowing plants to escape drought prone seasons, then drought
frequency should predict the distribution of life history strategies
across landscapes, and annual species should be more commonly associated
with drought prone regions than perennial species. Additionally, if
perenniality offers competative advantage in the absence of drought,
associations between life history and drought frequency should be
signficant in even if phylogenies are enriched for transitions from
annual to perenial life history strategy. Finally, if annual species
have adapted to escape predictably drought prone seasons, observations
of growing annual species (i.e.~occurring in forms other than seed)
should be rare during seasons when drought frequency is highest.
Phylogenetic relatedness can influence tests of associations between
species' traits and their environments (Felsenstein, 1985; Barrett
\emph{et al.}, 1996). Therefore we assessed the relationship between
life history distribution and drought frequency in a phylogenetic
context.

@@

All analyses were performed using R. All data and the source code to
produce this manuscript are available at
\url{https://github.com/greymonroe/heliophila}. Software used is listed
in the supplement.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
Africa including the Cape Floristic and Succulent Karoo Regions. These
are among the most botanically diverse environments on Earth and the
\emph{Heliophila} species occurring there are considered to make up the
most diverse genus of the family Brassicaceae (Mummenhoff \emph{et al.},
2005; Mandáková \emph{et al.}, 2012). This genus includes both perennial
(n =21) and annual (n = 21) species and this change in life history
strategy has likely arisen multiple independent times (Appel \&
Al-Shehbaz, 1997; Mummenhoff \emph{et al.}, 2005). Furthermore, the fine
scale climatic heterogeneity of Southern Africa is ideal for studying
the distribution of traits in relation to environmental parameters
(Sayre \emph{et al.}, 2013). We used life histories reported by
Mummenhoff \emph{et al.} (2005), grouping species with annual or
perennial life histories. Perenniality was defined based any form of
perennial life history (e.g., herbs, shrubs, mixed, etc). Because the
nature of species reported with mixed traits were unknown
(i.e.~plasticity vs.~genetic variation), we classified these species
here as perennial since they can maintain vegetative growth after
reproduction at least to some capacity.

@@

To characterize the distributions of annual and perennial
\emph{Heliophila} species, all (8670) records for the genus
\emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018 (GBIF, 2018). herbaria
records records such as these provide a rich data sources to
characterize the geographical distributions of species (Thiers, 2016;
Willis \emph{et al.}, 2017; Lang \emph{et al.}, 2018). And as they
become digitized (Soltis, 2017), these collections have been used to
study relationships between trait distributions, geography, and climate
(Davis \emph{et al.}, 2015; Stropp \emph{et al.}, 2016; Wolf \emph{et
al.}, 2016; Václavı'k \emph{et al.}, 2017).

@@

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak \emph{et al.}, 2015). From an ecological perspective,
droughts are best defined as episodes of plant stress caused by elevated
aridity (Passioura, 1996). Thus remote sensing offers the additional
benefit for studying drought as an agent of natural selection because
plant stress caused by drought can be observed from space (Kogan,
1995a). The remotely sensed Vegetative Health Index (VHI) is one such
metric, which detects landscape scale reductions in plant cover and
temperature conditions characteristic of drought (Kogan, 2001).
Generated from data collected by NOAA AVHRR satellites since 1981, the
VHI is a composite index combining Normalized Difference Vegetation
Index (NDVI) derived quantification of vegetative stress (Vegetative
Condition Index - VCI) with temperature stress indicated by anomalies in
thermal spectra (Temperature Condition Index - TCI). These indicies were
developed to create an unbiased quantification of drought across
ecosystem types. The VHI of year \(y\) during week \(w\) of \([1,52]\)
at pixel \(i\) is derived from the following equations, where \(n\) is
the number of years observed.

@@

\[TCI_{y,w,i} = 100\frac{T_{max,w,i}-T_{y,w,i}}{T_{max,w,i} - T_{min,w,i}}\]
Similarly, low TCI values indicate episdoes of high thermal stress shown
to be negatively correlated with precipition and soil moisture
(AghaKouchak \emph{et al.}, 2015).

\[VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})\] By combining VCI
and TCI, the VHI is distinguishes drought from other forms of vegetative
stress (Kogan, 1995b). The use of the VHI to detect drought has been
validated globally and across ecosystem types (AghaKouchak \emph{et
al.}, 2015), including in southern Africa, the focal region of this
study (Figure \ref{fig:mapsdroughtexamples}). To date, the VHI has most
often been applied for evaluating drought risk for agricultural research
(e.g., Rojas \emph{et al.}, 2011; Kogan \emph{et al.}, 2016). But it
also presents a tool to study seasonal patterns in the frequency of
drought across environments and to test hypotheses about the effect of
drought on ecological and evolutionary processes (Kerr \& Ostrovsky,
2003). As such, the VHI has been applied recently to study drought
related ecology of natural species and proven useful for predicting
intraspecific variation in drought tolerance traits and genes (Mojica
\emph{et al.}, 2016; Dittberner \emph{et al.}, 2018; Monroe \emph{et
al.}, 2018b). Here, we accessed VHI data at \(16km^2\) resolution from
1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species across their native range of
southern Africa.

@@

To characterize drought regimens across the distributions of annual and
perennial species of \emph{Heliophila}, we calculated drought during
different seasons at the location of observations for \emph{Heliophila}
records using the VHI. Specifically, we created global maps of the
frequencies of observing drought conditions between years
(VHI\textless{}40, NOAA) during the winter (quarter surrounding winter
solstice), spring (quarter surrounding spring equinox), summer (quarter
surrounding summer solstice) and fall (quarter surrounding fall equinox)
from 1981 to 2015. From these maps, the drought frequency during the
winter, spring, summer, and fall were extracted for the locations of all
GBIF records.

@@

Outgroup ( \emph{Aethionema grandiflorum, Alliaria petiolata, Cardamine
matthioli, Chamira circaeoides}, and \emph{Rorippa amphibia}) and
ingroup \emph{Heliophila} ITS I and II sequences were aligned using
MAFFT (Katoh \emph{et al.}, 2002) with strategy G-INS-I, offset value
0.1, and all other options set as default. The \(GTR + \Gamma\) model of
nucleotide substitution was determined to best fit the data based on AIC
using jModelTest2 (Guindon \& Gascuel, 2003; Darriba \emph{et al.},
2012). A maximum clade credibility tree with branch lengths as relative
time was estimated by summarizing data from six runs of 100,000,000
generations of Bayesian Markov chain Monte Carlo conducted in BEAST 2
(Bouckaert \emph{et al.}, 2014). Model selection and phylogenetic
analyses were conducted through the CIPRES Science Gateway (Miller
\emph{et al.}, 2010). Ancestral state estimation was performed in R
using the package phytools (Revell, 2012) to genearte 10,000 stochastic
character maps simulated under an equal rates model of character
evolution for the trait life habit (annual or perennial).

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought across all occurrence records of between annual and
perennial species by t-tests. Independent tests were performed to
compare drought frequency between annuals and perennials during each
season. We also tested for a relationship between drought frequency and
life history, season, and their interaction by an analysis of variance
while including species as a random effect. To account for variation in
the number of occurrence records per species, we next calculated the
mean drought frequency during the winter, spring, summer and fall for
each species. Because shared evolutionary history of closely related
species can lead to spurious associations between traits and
environments (Felsenstein, 1985), we tested for a relationship between
life history strategy and drought frequency while controlling for
phylogeny using phylogenetic logistic regression (Ives \& Garland,
2010). This statistical approach is designed to control for the
confounding effects of geographic relationships explained by demographic
history rather than processes of natural selection.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/phylogeny} \caption{Species and examples of herbaria specimens of
\emph{Heliophila} (a) Phylogeny and life history strategies of species
studied. Orange circles at branch tips mark annual species and blue
circles mark perennial species. At internal nodes, pie charts indicate
the estimated posterior probability of being annual versus perennial.
Example herbaria specimens accessed via GBIF of (a) \emph{H. minima},
(b) \emph{H. deserticola}, (c) \emph{H. coronopifolia} and (d) \emph{H.
ephemera}. Images (a,c,d) courtesy of The Bavarian Natural History
Collections (CC BY-SA 4.0) and (b) The London Natural History Museum (CC
BY 4.0). Links to images are found in the supplement.}\label{fig:phylogeny}
\end{figure}

>>>>

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Drought regimens predict life history strategies in Heliophila},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn G. Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, herbaria records, Heliophila, life history
evolution, phylogeography, remote sensing},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Drought regimens predict life history strategies in \emph{Heliophila}}
    \author{J. Grey Monroe\textsuperscript{1,2}, Brian Gill\textsuperscript{3},
Kathryn G. Turner\textsuperscript{4}, \& John K.
McKay\textsuperscript{2}}
    \date{}

\shorttitle{Drought and life history}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO 80521, USA\\\textsuperscript{2} College of Agriculture, Colorado State University, Fort Collins, CO 80521, USA\\\textsuperscript{3} Institute for Environment and Society, Brown University, Providence, RI 02912, USA\\\textsuperscript{4} Biology Department, Pennsylvania State University, State College, PA 16802, USA}
\keywords{drought adaptation, herbaria records, Heliophila, life history evolution, phylogeography, remote sensing }
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

Explaining variation in life history strategies is an enduring goal of
evolutionary biology and ecology. Early theory predicted that for
plants, annual and perennial life histories reflect adaptation to
environments that experience alternative drought regimens. Nevertheless,
empirical support for this hypothesis from comparative analyses remains
lacking.

@@

Here, we test classic life history theory in \textit{Heliophila}
(Brassicaceae), a diverse genus of flowering plants native to Africa,
controlling for phylogeny and integrating 34 years of satellite-based
drought detection with 2,192 herbaria occurrence records.

@@

We find that the common ancestor of \textit{Heliophila} species was
likely an annual, and that perenniality and annuality have repeatedly
evolved, eight and five times, respectively. By comparing historical
drought regimens, we show that annuals occur in environments where
droughts are significantly more frequent than perennial species. We also
provide evidence that annual plants adapt to predictable drought
regimens by escaping drought prone seasons as seeds.

@@

These results yield compelling support for longstanding theoretical
predictions by revealing the importance of drought frequency and
predictability to explain plant life history. More broadly, this work
highlights scalable approaches integrating herbaria records and remote
sensing to address outstanding questions in evolutionary ecology.

}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with some species
completing their life cycle in a number of weeks to others that live for
thousands of years (Brown, 1996). Along this continuum in angiosperms an
important division exists distinguishing annuals which complete their
seed to seed life cycle within a single calendar year from perennials
which can persist over multiple years. Annual plants do not need to
survive through the full range of seasonal environmental variation and
spend at least some portion of the year as a seed where they are
relatively protected from environmental stress. In contrast, perennial
plants can continue vegetative growth over multiple years and must
survive conditions experienced during all seasons but can also benefit
from competitive advantages and, if iteroparous, multiple bouts of
reproduction. These represent fundamentally different life history
strategies and predicting their occurrence is important for community,
ecosystem, and agricultural ecology. However, the environmental factors
that explain their evolution and distributions remain empirically
unresolved (Friedman \& Rubin, 2015).

Classical theory predicts shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Stearns, 1992; Franco \&
Silvertown, 1996). Because lack of water is perhaps the greatest threat
to survival during vegetative or reproductive growth in plants, this
theory has been extended to the hypothesis that annuality is adaptive
when it allows plants to escape drought (Schaffer \& Gadgil, 1975).
Indeed, adaptation to drought, defined as episodes of increased aridity
causing plant sress (Passioura, 1996), is often invoked as an
explanation for the success of annual species. And while a few cases are
cited where annuality appears to be more common in environments with
greater aridity (Stebbins Jr, 1952; Morishima \emph{et al.}, 1984), this
hypothesis has yet to be supported while controlling for the effect of
common ancestry (phylogeny) on life habit. In one previous study where
this question was addressed phylogenetically, (Evans \emph{et al.},
2005) annuals were not found to be associated with environments that
experience more drought. This could be explained by the relatively small
number of species studied and the reliance on a limited number of
weather stations to characterize environments, highlighting the need to
develop more scalable methods to study the geographic distributions of
traits such as life history. Thus, in this study we leverage thousands
of herbaria specimens among dozens of species and high-resolution remote
sensing to study the distributions and environmental factors potentially
driving the evolution and distribution of life history.

It is also critical to consider another dimension of drought adaptation:
the expectation that annuality is most adaptive when droughts are not
only frequent but also predictable. That is, when the frequency of
drought is particularly high during certain seasons. Such predictability
is important for selection to favor and escape strategy during those
seasons which are particularly drought prone. While there has been at
least one example of annuality associated with environments
qualitatively classified as \enquote{predictable} in a general sense
(Datson \emph{et al.}, 2008), the seasonal predictability of drought
experienced by annuals has yet to be rigorously studied. As such,
further empirical work is needed to support the model of annuality as a
mechanism of drought adaptation via escape from drought prone seasons.
Here we study herbarium collection dates to ask whether annuals indeed
exhibit evidence of an escape strategy from seasons with elevated
drought frequency.

In addition to drought escape in annuals as a mechanism of adaptation to
frequent and predictable droughts, droughts may be necessary for the
success of annuals more generally by acting as episodes of disturbance
that provide opportunities for annuals to establish and compete with
sympatric perennial species. Indeed, there is evidence that perennials
dominate in environments where disturbance events are infrequent (Rees
\& Long, 1992; Corbin \& D'Antonio, 2004; Clary, 2012). The resulting
prediction from this hypothesis is that in the absence of frequent
drought, perenniality should evolve. However, little is known about this
component of life history evolution because previous work has almost
entirely focused on the origins of annuality rather than perenniality
(Friedman \& Rubin, 2015). This highlights the need to study taxa which
have seen transitions from annual to perennial life histories as well.

Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses within the African endemic mustard genus,
\emph{Heliophila} L. (Brassicaceae). If annuality is an adaptive
strategy allowing plants to escape drought prone seasons, then drought
frequency should predict the distribution of life history strategies
across landscapes, and annual species should be more commonly associated
with drought prone regions than perennial species. Additionally, if
perenniality offers competitive advantage in the absence of drought,
associations between life history and drought frequency should be
significant when phylogenies include transitions from annual to
perennial life history strategy. Finally, if annual species have adapted
to escape predictably drought prone seasons, observations of growing
annual species (i.e.~occurring in forms other than seed) should be rare
during seasons when drought frequency is highest. Phylogenetic
relatedness can influence tests of associations between species' traits
and their environments by confounding common environments caused by
selection from common environments caused my ancestry. (Felsenstein,
1985; Barrett \emph{et al.}, 1996). Therefore, we assessed the
relationship between life history distribution and drought frequency
while controlling for phylogeny.

@@

All analyses were performed using R. All data and the source code to
produce this manuscript are available at
\url{https://github.com/greymonroe/heliophila}.

@@

\emph{Heliophila} is a genus of flowering plants endemic to southern
Africa including the Cape Floristic and Succulent Karoo Regions. These
are among the most botanically diverse environments on Earth and the
\emph{Heliophila} species occurring there are considered to be among the
most phenotypically diverse genera of the family Brassicaceae
(Mummenhoff \emph{et al.}, 2005; Mandáková \emph{et al.}, 2012). This
genus includes both perennial and annual species, and transitions
between life history strategy may have occurred multiple independent
times (Appel \& Al-Shehbaz, 1997; Mummenhoff \emph{et al.}, 2005).
Furthermore, the fine scale climatic heterogeneity of southern Africa is
ideal for studying the distribution of traits in relation to
environmental parameters (Sayre \emph{et al.}, 2013). We used life
histories reported by Mummenhoff \emph{et al.} (2005), grouping species
into annual or perennial life history categories. Perenniality was
defined as any form of perennial life history (e.g., herbs, shrubs,
mixed, etc). Because the nature of species reported with mixed traits
were unknown (i.e.~plasticity vs.~genetic variation), we classified
these species here as perennial since they can maintain vegetative
across multiple years at least to some capacity.

@@

To characterize the distributions of annual (studied here, n = 21) and
perennial (studied here, n = 21) \emph{Heliophila} species, all (8670)
records for the genus \emph{Heliophila} were downloaded from the Global
Biodiversity Information Facility (gbif.org) on July 21, 2018 (GBIF,
2018). Herbaria records such as these provide a rich data sources to
characterize the geographical distributions of species (Thiers, 2016;
Willis \emph{et al.}, 2017; Lang \emph{et al.}, 2018). And as they
become digitized (Soltis, 2017), herbaria collections have been used to
study relationships between trait distributions, geography, and climate
(Davis \emph{et al.}, 2015; Stropp \emph{et al.}, 2016; Wolf \emph{et
al.}, 2016; Václavı'k \emph{et al.}, 2017).

@@

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak \emph{et al.}, 2015). From an ecological perspective,
droughts are best defined as episodes of plant stress caused by elevated
aridity (Passioura, 1996). Thus remote sensing offers the additional
benefit for studying drought as an agent of natural selection because
plant stress caused by drought can be observed from space (Kogan,
1995a). The remotely sensed Vegetative Health Index (VHI) is one such
metric, which detects landscape scale reductions in plant cover and
temperature conditions characteristic of drought (Kogan, 2001).
Generated from data collected by NOAA AVHRR satellites since 1981, the
VHI is a composite index combining Normalized Difference Vegetation
Index (NDVI) derived quantification of vegetative stress (Vegetative
Condition Index - VCI) with temperature stress indicated by anomalies in
thermal spectra (Temperature Condition Index - TCI). These indices were
developed to create an unbiased quantification of drought across
ecosystem types. The VHI of year \(y\) during week \(w\) of \([1,52]\)
at pixel \(i\) is derived from the following equations, where \(n\) is
the number of years observed.

@@

\[TCI_{y,w,i} = 100\frac{T_{max,w,i}-T_{y,w,i}}{T_{max,w,i} - T_{min,w,i}}\]
Similarly, low TCI values indicate episdoes of high thermal stress shown
to be negatively correlated with precipitation and soil moisture
(AghaKouchak \emph{et al.}, 2015).

\[VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})\] By combining VCI
and TCI, the VHI distinguishes drought from other forms of vegetative
stress (Kogan, 1995b). The use of the VHI to detect drought has been
validated globally and across ecosystem types (AghaKouchak \emph{et
al.}, 2015), including in southern Africa, the focal region of this
study (e.g.~Figure \ref{fig:mapsdroughtexamples}). To date, the VHI has
most often been applied for evaluating drought risk for agricultural
research (e.g., Rojas \emph{et al.}, 2011; Kogan \emph{et al.}, 2016).
But it also presents a tool to study seasonal patterns in the frequency
of drought across environments and to test hypotheses about the effect
of drought on ecological and evolutionary processes (Kerr \& Ostrovsky,
2003). As such, the VHI has been applied recently to study drought
related ecology of natural species and proven useful for predicting
intraspecific variation in drought tolerance traits and genes (Mojica
\emph{et al.}, 2016; Dittberner \emph{et al.}, 2018; Monroe \emph{et
al.}, 2018b). Here, we accessed VHI data at \(16km^2\) resolution from
1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species across their native range of
southern Africa.

@@

To characterize drought regimens across the distributions of annual and
perennial species of \emph{Heliophila}, we calculated drought during
different seasons at the location of observations for \emph{Heliophila}
records using the VHI. Specifically, we created maps of the frequencies
of observing drought conditions between years (VHI\textless{}40, NOAA)
during the winter (quarter surrounding winter solstice), spring (quarter
surrounding spring equinox), summer (quarter surrounding summer
solstice) and fall (quarter surrounding fall equinox) from 1981 to 2015
across the range of \emph{Heliophila}. From these maps, the drought
frequency (the number of times drought is observed divided by the total
number of years, 34) during the winter, spring, summer, and fall were
extracted for the locations of all GBIF records.

@@

Outgroup ( \emph{Aethionema grandiflorum, Alliaria petiolata, Cardamine
matthioli, Chamira circaeoides}, and \emph{Rorippa amphibia}) and
ingroup \emph{Heliophila} ITS I and II sequences were aligned using
MAFFT (Katoh \emph{et al.}, 2002) with strategy G-INS-I, offset value
0.1, and all other options set as default. The \(GTR + \Gamma\) model of
nucleotide substitution was determined to best fit the data based on AIC
using jModelTest2 (Guindon \& Gascuel, 2003; Darriba \emph{et al.},
2012). A maximum clade credibility tree with branch lengths as relative
time was estimated by summarizing data from six runs of 100,000,000
generations of Bayesian Markov chain Monte Carlo conducted in BEAST 2
(Bouckaert \emph{et al.}, 2014). Model selection and phylogenetic
analyses were conducted through the CIPRES Science Gateway (Miller
\emph{et al.}, 2010). Ancestral state estimation was performed in R
using the package phytools (Revell, 2012) to generate 10,000 stochastic
character maps simulated under an equal rates model of character
evolution for the trait life habit (annual or perennial).

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimens, we
tested the corresponding prediction that the observed distributions of
annual and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought across all occurrence records of between annual and
perennial species by t-tests. Independent tests were performed to
compare drought frequency between annuals and perennials during each
season. We also tested for a relationship between drought frequency and
life history, season, and their interaction by analysis of variance
while including species as a random effect. To account for variation in
the number of occurrence records per species, we next calculated the
mean drought frequency during the winter, spring, summer and fall for
each species. Because shared evolutionary history of closely related
species can lead to spurious associations between traits and
environments (Felsenstein, 1985), we tested for a relationship between
life history strategy and drought frequency while controlling for
phylogeny using phylogenetic logistic regression (Ives \& Garland,
2010). This statistical approach is designed to control for the
confounding effects of common ancestry's influence on demographic
features such as geospatial relationships when addressing hypotheses
about the role of natural selection on trait distributions.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/phylogeny} \caption{Species and examples of herbaria specimens of
\emph{Heliophila} (a) Phylogeny and life history strategies of species
studied. Orange circles at branch tips mark annual species and blue
circles mark perennial species. At internal nodes, pie charts indicate
the estimated posterior probability of being annual versus perennial.
Example herbaria specimens accessed via GBIF of (a) \emph{H. minima},
(b) \emph{H. deserticola}, (c) \emph{H. coronopifolia} and (d) \emph{H.
ephemera}. Images (b,d,e) courtesy of The Bavarian Natural History
Collections (CC BY-SA 4.0) and (c) The London Natural History Museum (CC
BY 4.0). Links to images are found in the supplement.}\label{fig:phylogeny}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/maps_boxplots} \caption{Locations of occurrence records of (a) annual and (b)
perennial \emph{Heliophila.} Drought frequency during the (c) winter,
(d) spring, (e) summer and (f) fall detected using the VHI. (g) Drought
frequencies during each season at the observation locations of annual
and perennial \emph{Heliophila} (Post-hoc contrasts annuals and
perennials from ANOVA, ** = p \textless{} 0.01).}\label{fig:mapsboxplots}
\end{figure}
